---
title: "Unit 5 - Moving Average (MA(q)) and ARMA(p,q) Models"
output:
  html_document:
    df_print: paged
---


# For Live Session

In preparation for the next live session, please complete the following. Be sure and submit your work to the "Unit 5: "For Live Session" Assignment" assignment on 2DS:

- Use Aic5 to assess the use of ARMA models for your time series.  Add your findings to the Google doc. 

For the breakout PowerPoint slides:

- Slide 1: Use Aic5 to assess the use of ARMA models in the Walmart data.

- Slide 2: Find ρ1 for the following model by hand.  Xt = at–.8at-1 + .5at–2.

- Slide 3: Represent the model as a GLP.

- Slide 4: Generate a realizations from an ARMA model. You pick p and q.  Include the ACF and spectral density. 

- Slide 5: Use AIC 5 to identify the top five quality models with respect to AIC for the cancelled flight data from the SWADelay.csv data set (column: arr_cancelled). Comment on which are AR, MA, and ARMA.

- Slide 6+: Provide a slide or slides for your Key Takeaways for Unit 5!

- Last slide: Provide a slide or slides for any questions or comments you have from this section. This could also include topics you would like to have specifically addressed in live session! There is no minimum or maximum here.



```{r library-imports}

library(tidyverse)
library(tswge)


```

# For Live Session

## Bikeshare Demand

```{r for-live-session-google-doc-1}

df_train <- read.csv(file="../../Unit 2/pre_live/data/train.csv", header = TRUE)

# df_train$datetime = as.POSIXct(df_train$datetime)

df_train$datetime = as.POSIXct(strptime(df_train$datetime, "%Y-%m-%d %H:%M:%S"))


df_train <- df_train %>%
  mutate(year = as.factor(format(datetime, format = "%Y")), 
         month = as.numeric(format(datetime, format = "%m")), 
         day = as.factor(format(datetime, format = "%d")),
         hour = as.factor(format(datetime, format = "%H")))

```


```{r for-live-session-google-doc-2}

df_dec_2012 <- df_train %>%
    filter(datetime >= "2012-12-01")
  
```

```{r for-live-session-google-doc-3}

aic5.wge(df_dec_2012$count)

```



## Walmart Data

```{r for-live-session-slide-1}

#Walmart Analysis 
# Walmart = read.csv(file.choose(),header = TRUE)
Walmart = read.csv(file="../../Unit 3/Walmart.csv", header = TRUE)


Store9Item50 = Walmart %>% filter(store == "9", item == "50")

plotts.wge(Store9Item50$sales)
plotts.sample.wge(Store9Item50$sales)

aic5.wge(Store9Item50$sales)

# parzen.wge(na.omit(Store9Item50$sales))
# parzen.wge(na.omit(Store8Item50$sales), trunc = 500)


# plotts.wge(SWA$weather_delay)
# plotts.sample.wge(SWA$weather_delay)

# aic5.wge(SWA$weather_delay)

```


## Rho 1 Calculation

### General Equation for rho 1

$$\begin{equation}
   \rho_1 = \frac{-\theta_1 + \theta_1\theta_2}{1 + \theta_1^2 + \theta_2^2}
\end{equation}$$

### Actual Equation

$$\begin{equation}
   X_{t} = a_{t} - .8X_{t-1} + .5X_{t-2}
\end{equation}$$

```{r for-live-session-slide-2}
# X_t = a_t -.8a_t-1 + .5a_t-2

theta1 = .8
theta2 = -.5

p0 = 1
p1 = NA

p1 = (-(theta1) + (theta1 * theta2)) / (1 + (theta1^2 + theta2^2))
p1

```



```{r for-live-session-slide-4}

# ARMA
plotts.true.wge(phi = c(.3, .9, .1, -.8075), theta = c(-.9, -.8, -.72))
# factor.wge(c(.3, .9, .1, -.8075))

```


## SWA Cacelled Flights

Use AIC 5 to identify the top five quality models with respect to AIC for the cancelled flight data from the SWADelay.csv data set (column: arr_cancelled). Comment on which are AR, MA, and ARMA.

```{r for-live-session-slide-5}

# aic.wge(jetA$Price, p = 1, q = 0 )$value

# plotts.ege(jetA$Price)

# aic5.wge(jetA$price)

# SWA Delay Analysis

# SWA = read.csv(file.choose(), header = TRUE)
SWA = read.csv(file="../SWADelay.csv", header = TRUE)

plotts.wge(SWA$arr_cancelled)
plotts.sample.wge(SWA$arr_cancelled)
 
aic5.wge(SWA$arr_cancelled)

```


# Asynchronous Material

## 5.2 MA(q): Properties and Characteristics, Part I


### 5.2.5 MA(1) Model


$$\begin{equation}
   X_{t} = a_{t} -.5X_{t-1}
\end{equation}$$

```{r 5.2.5}

# X_t = a_t -.5a_t-1


theta1 = .5

p0 = 1
p1 = NA
p2 = 0
p3 = 0


p1 = -(theta1) / (1 + theta1^2)
p1


```

#### Equation for rho 1

$$\begin{equation}
   \rho_1 = \frac{-\theta_1}{1 + \theta_1^2}
\end{equation}$$


### 5.2.7 MA(2) Model


$$\begin{equation}
   X_{t} = a_{t} -.5X_{t-1} + .9X_{t-2}
\end{equation}$$

```{r 5.2.7}

# X_t = a_t -.5a_t-1 + .9a_t-2


theta1 = .5
theta2 = -.9

p0 = 1
p1 = NA
p2 = NA
p3 = 0


p1 = (-(theta1) + (theta1 * theta2)) / (1 + (theta1^2 + theta2^2))
p1


p2 = -(theta2) / (1 + (theta1^2 + theta2^2))
p2


```
#### Equation for rho 1

$$\begin{equation}
   \rho_1 = \frac{-\theta_1 + \theta_1\theta_2}{1 + \theta_1^2 + \theta_2^2}
\end{equation}$$


#### Equation for rho 2

$$\begin{equation}
   \rho_2 = \frac{-\theta_2}{1 + \theta_1^2 + \theta_2^2}
\end{equation}$$


## 5.3 MA(q): Properties and Characteristics, Part II

### 5.3.1 tswge


```{r 5.3.1.1}

# Generating MA(1) data

# gen.arma.wge(n=100, theta=.99)
# gen.arma.wge(n=100, theta=-.99)
# gen.arma.wge(n=100, theta=.99, sn=5)
# gen.arma.wge(n=100, theta=-.99, sn=5)
plotts.true.wge(theta=c(.99))


```


```{r 5.3.1.2}

# Generating MA(2) data from Example 2

# gen.arma.wge(n=100, theta=c(.9,-.4))
plotts.true.wge(theta=c(.9,-.4))


```

### 5.3.4 tswge


```{r 5.3.4.1}

# AR(2) X_t -1.1X_t-1 + .9X_t-2 = a_t

plotts.true.wge(phi=c(1.1,-.9))

```

```{r 5.3.4.2}

# MA(2) X_t = a_t -1.1a_t-1 + .9a_t-2 

plotts.true.wge(theta=c(1.1,-.9))

```


### 5.4.3 MA(q) Invertablity

$$\begin{equation}
   X_{t} = a_{t} - 1.6X_{t-1} + .9X_{t-2}
\end{equation}$$

```{r 5.4.3.1}


# MA(2) X_t = a_t -1.6a_t-1 + .9a_t-2 

factor.wge(phi=c(1.6,-.9))

### Abs Recip < 1 (and therefore outside of the unit circle which is invertable)

```
$$\begin{equation}
   X_{t} = a_{t} + 1.6X_{t-1} - .9X_{t-2}
\end{equation}$$


```{r 5.4.3.2}


# MA(2) X_t = a_t + 1.6a_t-1 - .9a_t-2 

# factor.wge(phi=c(1.6,.9))

factor.wge(phi=c(-.1,.3))

### Abs Recip > 1  (at least one root is inside of the unit circle and therefore not invertable)

```

## 5.5 ARMA(p,q), Properties and Characteristics


### ARMA(p,q) | Cancellation

These two models are the same due to the factors cancelling each other out.

```{r 5.5.4}

plotts.true.wge(250, phi = c(1.3,-.4), theta=c(.8))

# plotts.true.wge(250, phi = c(.5)


# 5.5.5
# plotts.true.wge(250, phi = c(-.1,-.82, .16), theta=c(-.2))
# 
# plotts.true.wge(250, phi = c(.1, -.8))

# 5.5.6
# plotts.true.wge(250, phi = c(1.95,-1.9), theta=c(1.95, -1.9))
# 
# plotts.true.wge(250, phi = c(1.95))

```


```{r 5.5.8}

# AR Factors

# plotts.true.wge(phi = c(.3, .9, .1, -.8075))
# factor.wge(c(.3, .9, .1, -.8075))


# MA Factors

# plotts.true.wge(theta = c(-.9, -.8, -.72))
# factor.wge(c(-.9, -.8, -.72))


# ARMA

plotts.true.wge(phi = c(.3, .9, .1, -.8075), theta = c(-.9, -.8, -.72))
# factor.wge(c(.3, .9, .1, -.8075))


# 5.5.10
# plotts.true.wge(phi = c(.1, -.5, .7), theta = c(.72, -.8))

```


## 5.6 ARMA(p,q), Examples


Akaike's Information Criterion (AIC)

Given a set of data, the AIC is used to evaluate and compare the quality of models.

The model with the lowest AIC is preferred.


### 5.6.4


```{r 5.6.4}

# aic.wge(jetA$Price, p = 1, q = 0 )$value

# plotts.ege(jetA$Price)

# aic5.wge(jetA$price)


# SWA Delay Analysis

# SWA = read.csv(file.choose(), header = TRUE)
SWA = read.csv(file="../SWADelay.csv", header = TRUE)

plotts.wge(SWA$weather_delay)
plotts.sample.wge(SWA$weather_delay)

aic5.wge(SWA$weather_delay)

```




## 5.7 ARMA(p,q), Psi Weights

### Example



```{r 5.7.2}

# psi-weights for simple MA(1) model X(t)=(1 - .8B)a(t)
psi.weights.wge(theta = .8, lag.max = 5)

# psi-weights for simple AR(1) model  X(t)=(1 - .8B)a(t)
psi.weights.wge(phi = .8, lag.max = 5)  # note that psi(j) = .8j

# psi-weights for ARMA(2,1) model (1 - 1.2B + .6B2)X(t) = (1 - .5B)a(t)
psi.weights.wge(phi=c(1.2, -.6), theta=c(.5), lag.max=5)

# 5.7.3
psi.weights.wge(phi = c(1.95, -1.9), lag.max = 5)

```

