---
title: "Time Series Analysis for IT Hardware Inventory Forecasts"
author: "Chance Robinson"
date: 10/26/2021
lang: en-US
class: man
# figsintext: true
numbersections: true
encoding: UTF-8
bibliography: references.bib
biblio-style: apalike
output:
  bookdown::pdf_document2:
     citation_package: natbib
     keep_tex: true
     toc: true
header-includes:
   - \usepackage{amsmath}
   - \usepackage[utf8]{inputenc}
   - \usepackage[T1]{fontenc}
   - \usepackage{setspace}
   - \usepackage{hyperref}
   - \onehalfspacing
   - \newcommand\numberthis{\addtocounter{equation}{1}\tag{\theequation}}
editor_options:
  chunk_output_type: console
---

\newpage


```{r library-imports, results='hide', message=FALSE, include=FALSE, echo=FALSE}

# Time Series Analysis
library(tseries)
library(forecast)
library(TSstudio)
library(tswge)
library(orcutt)

# Date Manipulation
library(lubridate)

# Metrics
library(MLmetrics)

# Misc
library(tidyverse)
library(kableExtra)
library(knitr)


```


```{r load-data, results='hide', message=FALSE, include=FALSE, echo=FALSE}

setwd('.')

# df_data <- read.csv(file="./Project/it_purchases/data/sc_req_item.csv", header = TRUE)
df_data <- read.csv(file="../data/sc_req_item.csv", header = TRUE)

# df_optiplex$opened_at = as.POSIXct(strptime(df_optiplex$opened_at, "%Y-%m-%d %H:%M:%SS"))

df_data$opened = floor_date(mdy_hm(df_data$opened_at), "day")

df_data <- df_data %>%
  mutate(year = as.factor(format(opened, format = "%Y")),
         month = as.factor(format(opened, format = "%m")))

df_optiplex <- df_data %>%
    filter(cat_item == "Dell Optiplex 7000 Series")
    # filter(cat_item == "Dell Latitude 14 5000 Series")



```


# Dell Optiplex

## Original Data

Includes only requests created after July, 28th 2018 and before October 21st, 2021 that are in a state of Work in Progress, Pending Verification or Closed Complete.  This is due to the fact that a high percentage of orders are cancelled before being fully approved in PeopleSoft, which should not necessarily be included for inventory estimation purposes.

### Daily

The average amount of products requested for the period was __`r round(mean(df_optiplex$quantity), digits=2)`__.  


```{r optiplex-line-chart-by-day-original, message=FALSE, echo=FALSE}

ggplot(data=df_optiplex, aes(x = opened, y = quantity)) +
      geom_line() + 
      ggtitle("Daily Optiplex Purchases") +
      labs(x = "Daily", y = "Quantity Purchased")

```

\newpage

```{r optiplex-bar-chart-by-quantity-original, results='hide', message=FALSE, include=FALSE, echo=FALSE}

df_optiplex %>%
  remove_missing(na.rm = TRUE) %>%
  group_by(quantity) %>%
  # filter(n() < 10) %>%
  ggplot(aes(x = quantity, fill=..count..)) + 
  geom_bar() +
  geom_text(stat='count', aes(label=..count..), vjust=-1) +
  ggtitle("Optiplex Purchases by Quantity") +
  labs(x = "Quantity", y = "Count") +
  theme(plot.title = element_text(hjust = 0.5)) + lims(y=c(0,2000)) 

```

#### Histogram of Counts by Quantity

- There are a few purchases where a single order had more than 5 items (i.e..the quantity)  
- We will assume that these would normally go through the normal PeopleSoft purhcasing lifecycle.



```{r optiplex-histogram-by-quantity-original, message=FALSE, echo=FALSE}
df_optiplex %>%
  ggplot(aes(quantity)) +
  geom_histogram(bins = 25) +
  ggtitle("Histogram of Optiplex Purchases by Quantity") +
  labs(x = "Quantity", y = "Count") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))

```

\newpage

### Weekly


```{r optiplex-counts-by-week-original, results='hide', message=FALSE, include=FALSE, echo=FALSE}

df_optiplex_by_week_orig <- df_optiplex %>% 
    group_by(week = lubridate::floor_date(opened, "week")) %>%
    summarize(summary_variable = sum(quantity))

```

The average amount of products requested for the period was __`r round(mean(df_optiplex_by_week_orig$summary_variable), digits=2)`__. 

```{r optiplex-line-chart-by-week-original, message=FALSE, echo=FALSE}

ggplot(data=df_optiplex_by_week_orig, aes(x = week, y = summary_variable)) +
      geom_hline(yintercept = mean(df_optiplex_by_week_orig$summary_variable), color="blue") + 
      geom_line() +
      ggtitle("Weekly Optiplex Purchases") +
      labs(x = "Week", y = "Quantity Purchased") 

```

\newpage

### Monthly

```{r optiplex-counts-by-month-original, results='hide', message=FALSE, include=FALSE, echo=FALSE}

df_optiplex_by_month_orig <- df_optiplex %>% 
    group_by(month = lubridate::floor_date(opened, "month")) %>%
    summarize(summary_variable = sum(quantity))

```

The average amount of products requested for the period was __`r round(mean(df_optiplex_by_month_orig$summary_variable), digits=2)`__. 

```{r optiplex-line-chart-by-month-original, message=FALSE, echo=FALSE}

ggplot(data=df_optiplex_by_month_orig, aes(x = month, y = summary_variable)) +
      geom_hline(yintercept = mean(df_optiplex_by_month_orig$summary_variable),color="blue") + 
      geom_line() +
      ggtitle("Monthly Optiplex Purchases") +
      labs(x = "Month", y = "Quantity Purchased")

```


\newpage


### Quarterly

```{r optiplex-counts-by-quarter-original, results='hide', message=FALSE, include=FALSE, echo=FALSE}

# https://stackoverflow.com/questions/33221425/how-do-i-group-my-date-variable-into-month-year-in-r/33221885

df_optiplex_by_quarter_orig <- df_optiplex %>% 
    group_by(quarter = lubridate::floor_date(opened, "quarter")) %>%
    summarize(summary_variable = sum(quantity))

```

The average amount of products requested for the period was __`r round(mean(df_optiplex_by_quarter_orig$summary_variable), digits=2)`__. 


```{r optiplex-line-chart-by-quarter-original, message=FALSE, echo=FALSE}

ggplot(data=df_optiplex_by_quarter_orig, aes(x = quarter, y = summary_variable)) +
      geom_hline(yintercept = mean(df_optiplex_by_quarter_orig$summary_variable), color="blue") + 
      geom_line() +
      ggtitle("Quarterly Optiplex Purchases") +
      labs(x = "Quarter", y = "Quantity Purchased") 

```

\newpage

### Quarterly Pivot Table

Note that the first and last quarters are incomplete and do not represent full quarters.

```{r optiplex-pivot-table-by-quarter-original, results='asis', message=FALSE, echo=FALSE}

df_optiplex_by_quarter_orig %>% 
  # kable(format = "html", booktabs = T)
  kable(format = "latex")
  # kable_styling(bootstrap_options = c("striped", "condensed", "center"))

```


\newpage

## Filtered Data

- Excluding requested items that had a quantity of less than five.
- Only looking a records created at or after July 5th, 2021 and before June 27, 2021 as the recent trend behaviour seems to behave differently to that of prior years.
- Also, there is a higher percentage of items yet to be processed in PeopleSoft which would skew our numbers if the more recent purhcases were to be included.


```{r optiplex-filter-records, results='hide', message=FALSE, include=FALSE, echo=FALSE}

df_optiplex <- df_optiplex %>%
  filter(quantity < 5) %>%
  filter(opened >= "2020-07-05")  %>%
  filter(opened < "2021-06-27")
  # filter(opened < "2021-07-01")
  
```

### Weekly

```{r optiplex-counts-by-week, results='hide', message=FALSE, include=FALSE, echo=FALSE}

df_optiplex_by_week <- df_optiplex %>% 
    group_by(week = lubridate::floor_date(opened, "week")) %>%
    summarize(summary_variable = sum(quantity))

```

The average amount of products requested for the period was __`r round(mean(df_optiplex_by_week$summary_variable), digits=2)`__. 

```{r optiplex-line-chart-by-week, message=FALSE, echo=FALSE}

ggplot(data=df_optiplex_by_week, aes(x = week, y = summary_variable)) +
      geom_hline(yintercept = mean(df_optiplex_by_week$summary_variable), color="blue") + 
      geom_line() +
      ggtitle("Weekly Optiplex Purchases") +
      labs(x = "Week", y = "Quantity Purchased") 

```

\newpage 

### Monthly

```{r optiplex-counts-by-month, results='hide', message=FALSE, include=FALSE, echo=FALSE}

df_optiplex_by_month <- df_optiplex %>% 
    group_by(month = lubridate::floor_date(opened, "month")) %>%
    summarize(summary_variable = sum(quantity))

```

The average amount of products requested for the period was __`r round(mean(df_optiplex_by_month$summary_variable), digits=2)`__. 

```{r optiplex-line-chart-by-month, message=FALSE, echo=FALSE}

ggplot(data=df_optiplex_by_month, aes(x = month, y = summary_variable)) +
      geom_hline(yintercept = mean(df_optiplex_by_month$summary_variable),color="blue") + 
      geom_line() +
      ggtitle("Monthly Optiplex Purchases") +
      labs(x = "Month", y = "Quantity Purchased")

```

\newpage 

### Quarterly

```{r optiplex-counts-by-quarter, results='hide', message=FALSE, include=FALSE, echo=FALSE}

df_optiplex_by_quarter <- df_optiplex %>% 
    group_by(quarter = lubridate::floor_date(opened, "quarter")) %>%
    summarize(summary_variable = sum(quantity))

```

The average amount of products requested for the period was __`r round(mean(df_optiplex_by_quarter$summary_variable), digits=2)`__. 

```{r optiplex-line-chart-by-quarter, message=FALSE, echo=FALSE}

ggplot(data=df_optiplex_by_quarter, aes(x = quarter, y = summary_variable)) +
      geom_hline(yintercept = mean(df_optiplex_by_quarter$summary_variable), color="blue") + 
      geom_line() +
      ggtitle("Quarterly Optiplex Purchases") +
      labs(x = "Quarter", y = "Quantity Purchased") 

```

\newpage 

#### Bar Chart by Quantity Purchased

- Note that all purchases with quantities exceeding 4 have been filtered.

```{r optiplex-bar-chart, message=FALSE, echo=FALSE}

df_optiplex %>%
  remove_missing(na.rm = TRUE) %>%
  group_by(quantity) %>%
  # filter(n() < 10) %>%
  ggplot(aes(x = quantity, fill=..count..)) + 
  geom_bar() +
  geom_text(stat='count', aes(label=..count..), vjust=-1) +
  ggtitle("Optiplex Purchases by Quantity") +
  labs(x = "Quantity", y = "Count") +
  theme(plot.title = element_text(hjust = 0.5)) + lims(y=c(0,550)) 


```



```{r optiplex-factor, results='hide', message=FALSE, include=FALSE, echo=FALSE}

factor.wge(phi = c(rep(0,11), 1))

```


```{r optiplex-estimate, results='hide', message=FALSE, include=FALSE, echo=FALSE}

est.ar.wge(df_optiplex_by_week$summary_variable, p = 15, type = "burg")

```


```{r optiplex-plotts-sample, results='hide', message=FALSE, include=FALSE, echo=FALSE}

plotts.sample.wge(df_optiplex_by_week$summary_variable)

```

```{r optiplex-log, results='hide', message=FALSE, include=FALSE, echo=FALSE}

# df_optiplex_by_week$summary_variable <- log(df_optiplex_by_week$summary_variable)

```


```{r optiplex-difference, results='hide', message=FALSE, include=FALSE, echo=FALSE}

d1 = artrans.wge(df_optiplex_by_week$summary_variable, phi.tr = 1)

d2 = artrans.wge(d1, phi.tr = 1)

```



```{r optiplex-aic-5, results='hide', message=FALSE, include=FALSE, echo=FALSE}

aic5.wge(df_optiplex_by_week$summary_variable,  p = 0:15, q = 0:2, type = "aic")


```




```{r optiplex-tswge-estimate, results='hide', message=FALSE, include=FALSE, echo=FALSE}

e.optiplex = est.arma.wge(df_optiplex_by_week$summary_variable, p=13, q=1)
mean(df_optiplex_by_week$summary_variable)

# e = e + mean(df_optiplex_by_week$summary_variable)

```

```{r optiplex-parzen, results='hide', message=FALSE, include=FALSE, echo=FALSE}

plotts.parzen.wge(df_optiplex_by_week$summary_variable)

```

\newpage 

## Model Performance

The chart below represents predictions made on observations that we already had actual counts for.  This will allow us to guage performance for future predictions.

### Predictions Plot

```{r optiplex-tswge-forecast,  message=FALSE, echo=FALSE}

f.optiplex = fore.aruma.wge(df_optiplex_by_week$summary_variable, phi = e.optiplex$phi, theta = e.optiplex$theta, n.ahead = 12, limits = FALSE, lastn=TRUE) 


```

### Predictions ASE

Average Squared Error

```{r optiplex-ase, results='hide', message=FALSE, include=FALSE, echo=FALSE}

seasonal_ase_optiplex = mean((df_optiplex_by_week$summary_variable[(51 - 12 + 1): 51] - f.optiplex$f)^2)
seasonal_ase_optiplex


```


We were able to achieve an ASE of __`r round(seasonal_ase_optiplex, digits=2)`__.  In this case, we overestimated the quarterly counts by about __`r round(seasonal_ase_optiplex)`__ items.

```{r optiplex-output, results='hide', message=FALSE, include=FALSE, echo=FALSE}

df_optiplex_out <- df_optiplex_by_week %>%
   select(week, summary_variable) %>%
   filter(week >= "2021-03-28")


df_optiplex_out$predictions = round(f.optiplex$f)

df_optiplex_out <- plyr::rename(df_optiplex_out, replace = c("summary_variable"="actual"))

df_optiplex_out <- df_optiplex_out %>%
  mutate(diff = (actual - predictions)) %>%
  mutate(abs.diff = abs(actual - predictions))


```

\newpage

## Forecast

### Actual Counts vs. Predicted

This table shows a comparison of the model's predictions vs. those of the known weekly counts from the prior 12 periods.



```{r optiplex-print-output, results='asis', message=FALSE, echo=FALSE}


df_optiplex_out %>% 
  # kable(format = "html", booktabs = T) %>% 
  kable(format = "latex", booktabs = T)
  

```


\newpage 

```{r optiplex-arima-model-1, results='hide', message=FALSE, include=FALSE, echo=FALSE}

arima.optiplex <- arima(df_optiplex_by_week$summary_variable, order=c(13, 0, 1))

```

### Future Predictions Plot


```{r optiplex-arima-model-2, results='hide', message=FALSE, include=FALSE, echo=FALSE}

forecast.optiplex <- forecast(arima.optiplex, h=12)

```


```{r optiplex-arima-model-3, message=FALSE, echo=FALSE}

plot(forecast.optiplex)
legend(x = "topleft", legend = c("Predicted"), col = c("blue"), lty = c(1, 1))

```

\newpage 

### Future Prediction Estimates Table

```{r optiplex-arima-model-4, results='asis', message=FALSE, echo=FALSE}

forecast.optiplex %>% 
  # kable(format = "html", booktabs = T) 
  kable(format = "latex", booktabs = T)
  


```


```{r optiplex-signal-plus-noise, results='hide', message=FALSE, include=FALSE, echo=FALSE}

# x = gen.sigplusnoise.wge(100, b0 = 0, b1 = 0, phi = .95, sn = 21, plot = FALSE)

# t = seq(1,51,1)
# 
# df = data.frame(x = df_optiplex_by_week$summary_variable, t=t)
# 
# fit = lm(df_optiplex_by_week$summary_variable~df_optiplex_by_week$week, data = df_optiplex_by_week)
#  
# summary(fit) # record whether it rejected or failed to reject. 

```

```{r optiplex-cochrane.orcut, results='hide', message=FALSE, include=FALSE, echo=FALSE}

# cfit = cochrane.orcutt(fit)
# summary(cfit)

```


```{r optiplex-scatter-plot, results='hide', message=FALSE, include=FALSE, echo=FALSE}

# df_optiplex_by_week %>%
#   # select(beer.abv, beer.ibu) %>%
#   remove_missing(na.rm = TRUE) %>%
#   ggplot(aes(x = week, y = summary_variable)) + 
#   ggtitle("Scatterplot of Quantity by Time") +
#   geom_point(alpha = 0.3) + 
#   labs(x = "Time", y = "Count") +
#   # scale_x_continuous(breaks = seq(from = 0, to = 45, by = 2)) + 
#   geom_smooth(method = 'lm')

```


