---
title: "Unit 7 - Forecasting"
output:
  html_document:
    df_print: paged
---

# For Live Session

```{r library-imports}

library(tidyverse)
library(tswge)

```


## Question 1 

1. Which model do you think is appropriate to use to forecast your time series? Why? Add this to the Google Doc as well as to your PowerPoint deck. 


```{r for-live-session-google-doc-1}

df_train <- read.csv(file="../../Unit 2/pre_live/data/train.csv", header = TRUE)

# df_train$datetime = as.POSIXct(df_train$datetime)

df_train$datetime = as.POSIXct(strptime(df_train$datetime, "%Y-%m-%d %H:%M:%S"))


df_train <- df_train %>%
  mutate(year = as.factor(format(datetime, format = "%Y")), 
         month = as.numeric(format(datetime, format = "%m")), 
         day = as.factor(format(datetime, format = "%d")),
         hour = as.factor(format(datetime, format = "%H")))

```


```{r for-live-session-google-doc-2}

df_dec_2012 <- df_train %>%
    filter(datetime >= "2012-12-01")
  
```

```{r for-live-session-google-doc-3}

plotts.sample.wge(df_dec_2012$count)

parzen.wge(df_dec_2012$count)

bike_diff = artrans.wge(df_dec_2012$count, c(rep(0,23), 1))  # Take out the (1-B^24)

aic5.wge(bike_diff) # Check the structure of the noise

parzen.wge(bike_diff)

```


## Question 2

2. Find the first 5 psi weights for the model below, and use them to find the half-width of the 95 percentile probability interval for the third forecast (Xhat(3)). Please show your work as well as a plot of the series and the first eight forecasts with probability intervals.

Hint: equations to help find sigma_hat_A and the formula for the confidence prediction interval can be found on pages 245 and 246


$$\begin{equation}
   (1 - .9B)(1 - .8B)X_{t} = a_{t}
\end{equation}$$

First 8 Observations:

X_1 = 5 
X_2 = 8
X_3 = 9
X_4 = 8
X_5 = 7
X_6 = 6
X_7 = 4
X_8 = 3


## Question 3


3.  Using the three models below and the Amtrak ridership data, which model has the smallest ASE in forecasting the next year (12 months)? Â Show your code in forecasting the 12 observations as well as in calculating the ASE.


**Model 1**

(1"+ 0.5511" ğµ"âˆ’ 0.1680" ğµ^2 "+0.0145" ğµ^3 "âˆ’ 0.0651" ğµ^4 "âˆ’ 0.1388" ğµ^5 "+ 0.2966" ğµ^6 "âˆ’ 0.1539" ğµ^7 "âˆ’ 0.1270" ğµ^8
"+0.1815" ğµ^9 "âˆ’ 0.0364" ğµ^10 "âˆ’ 0.1456" ğµ^11 "âˆ’ 0.6287" ğµ^12 "+ 0.3832" ğµ^13 "+ 0.0199" ğµ^14 "+ 0.1679" ğµ^15)ğ‘‹_ğ‘¡=ğ‘_ğ‘¡

phis:  c(0.5511, 0.1680, -0.0145, 0.0651, 0.1388, -0.2966, 0.1539, 0.1270, -0.1815, 0.0364, 0.1456, 0.6287, -0.3832, -0.0199, -0.1679) 


**Model 2**

(1"+ 0.027" 1ğµ"âˆ’0.7421" ğµ^2)(1âˆ’ğµ^12)ğ‘‹_ğ‘¡=ã€–(1"+ 0" .5845ğµ"âˆ’ " 0.3837ğµ^2)ğ‘ã€—_ğ‘¡

phis: c(-0.02709541,  0.74213105)
thetas: c(-0.5844596,  0.3836931)



**Model 3**

(1 âˆ’ .3069ğµ)(1âˆ’ğµ)(1âˆ’ğµ^12)ğ‘‹_ğ‘¡=ã€–(1"âˆ’ " 0.7432ğµ)ğ‘ã€—_ğ‘¡

phi: 0.306943
theta: 0.7431719



# Asynchronous Material

## 7.4 Forecasting with AR(p) Models

### AR(1) 

forecasts tend to approach the sample mean

```{r 7.4.2}

# (1 - 8B)(X_t - 25) = a_t

X_80 = 21.77
X_bar = 24.17

X_80_1 = 22.25
X_80_2 = 22.63
X_80_3 = 22.93

# X_t = **(1 - .8)mu** + .8X_t-1 + a_t

const = (1 - .8) * (X_bar)
round(const, 2)


X_80_4 = round(const + .8 * (X_80_3), 2)
X_80_4

X_80_5 = round(const + .8 * (X_80_4), 2)
X_80_5

X_80_6 = round(const + .8 * (X_80_5), 2)
X_80_6


```

### AR(2) 

forecasts tend to gradually in a sinusoidal pattern come back to the mean

$$\begin{equation}
   \hat {X_t}_0 (l) = \phi_1 \hat {X_t}_0 (l-1) + \phi_2 \hat {X_t}_0 (l-2) + \hat X(1 - \phi_1 - \phi_2)
\end{equation}$$



```{r 7.4.4}

# (1 - 1.6B + .8B^2)(X_t - 50) = a_t

X_75 = 23.4
X_74 = 27.7
X_bar = 29.4

# X_75_1 = 21.2
# X_75_2 = 21.1
# X_75_3 = 22.7

# X_bar(1 - phi1 - phi2)

phi1 = 1.6
phi2 = -.8

const = (X_bar) * (1 - phi1 - phi2)
round(const, 2)


X_75_1 = round( (1.6 * X_75) + (-.8 * X_74) + const, 2)
X_75_1 = 21.2

X_75_2 = round( (1.6 * X_75_1) + (-.8 * X_75) + const, 2)
X_75_2 = 21.1

X_75_3 = round( (1.6 * X_75_2) + (-.8 * X_75_1) + const, 2)
X_75_3 = 22.7

X_75_4 = round( (1.6 * X_75_3) + (-.8 * X_75_2) + const, 1)
X_75_4

X_75_5 = round( (1.6 * X_75_4) + (-.8 * X_75_3) + const, 1)
X_75_5

X_75_6 = round( (1.6 * X_75_5) + (-.8 * X_75_4) + const, 1)
X_75_6



```

### TSWGE

#### AR(1) Phi Positive

```{r 7.4.8.1, echo=FALSE, message=FALSE, results='hide'}

# AR(1) Phi Positive
data(fig6.1nf)
plotts.wge(fig6.1nf)

fore.arma.wge(fig6.1nf, phi = .8, n.ahead = 20, plot = TRUE, limits = FALSE)

```
#### AR(1) Phi Negative

```{r 7.4.8.2, echo=FALSE, message=FALSE, results='hide'}

# AR(1) Phi Negative

x1 = gen.arma.wge(100, phi = -.8)

fore.arma.wge(x1, phi = -.8, n.ahead = 20, plot = TRUE, limits = FALSE)

```


#### AR(2)

```{r 7.4.8.3, echo=FALSE, message=FALSE, results='hide'}

# AR(2)

x2 = gen.arma.wge(n=75, phi = c(1.6, -.8), sn=24)
x2 = x2 + 25
# plotts.wge(x2)

fore.arma.wge(x2, phi = c(1.6, -.8), n.ahead = 20, plot = TRUE, limits = FALSE)

```



### Eventual Forecast Function

Key Observations


1.  In both cases, the forecasts behave like the auto correlations.

- AR(1) case:  forecasts have exponential shape (although increasing instead of decreasing)
- AR(2) case:  forecasts look like a damped sinusoid (AR part has complex conjugate roots)


2.  The forecasts tend toward teh realization mean as number of steps ahead increases.



## 7.5 Forecasting with ARMA(p,q)


### Example


```{r 7.5.2.1, echo=FALSE, message=FALSE, results='hide'}

# AR(2)

x1 = gen.arma.wge(n=75, phi = c(1.6, -.8), sn=24)
fore.arma.wge(x1, phi = c(1.6, -.8), n.ahead = 20, plot = TRUE, limits = FALSE)


```

```{r 7.5.2.2, echo=FALSE, message=FALSE, results='hide'}

# ARMA(2,1)

x2 = gen.arma.wge(n=75, phi = c(1.6, -.8), theta = -.9, sn=24)
fore.arma.wge(x2, phi = c(1.6, -.8), theta = -.9, n.ahead = 20, plot = TRUE, limits = FALSE)


```



```{r 7.5.2.3 echo=FALSE, message=FALSE, results='hide'}

# ARMA(1,1)

x3 = gen.arma.wge(n=75, phi = c(.8), theta = -.9)
fore.arma.wge(x3, phi = c(.8), theta = -.9, n.ahead = 20, plot = TRUE, limits = FALSE)


```




### Canadian Lynx Data

```{r 7.5.3 echo=FALSE, message=FALSE, results='hide'}

# Canadian Lynx Data

data("llynx")
plotts.wge


# AR(4)
fore.arma.wge(llynx, c(1.3, -.7, .1, -.2), n.ahead = 20, limits = FALSE)

# ARMA(4,1)
fore.arma.wge(llynx, c(.7, .1, -.2, -.3), theta = -.6, n.ahead = 20, limits = FALSE)

# AR(11)
fore.arma.wge(llynx, c(1.1676, -0.5446, 0.2662, -0.3094, 0.1540, -0.1463, 0.0569, -0.0294, 0.1346, 0.2021, -0.3394), n.ahead = 20, limits = FALSE)


```

## 7.6 Probability Limits and ARMA Forecasts

### 7.6.4 Calculating Psi Weights with TSWGE


```{r 7.6.4 message=FALSE}

# psi_0 = 1, psi_1 = .7, psi_2 = .24, psi_3 = -.132, ...

psi.weights.wge(phi = c(1.2, -.6), theta = .5, lag.max = 5)


```

### 7.6.6 Probability Limits for ARMA Forecasts with an Example



```{r 7.6.6 message=FALSE}

sigma_hat = .8659

psi_0 = 1
psi_1 = .7
psi_2 = .24
psi_3 = -.132

l_1 = (1.96) * (sigma_hat) * psi_0
l_1

l_2 = (1.96) * (sigma_hat) * sqrt(1 + psi_1^2)
l_2


l_3 = (1.96) * (sigma_hat) * sqrt(1 + psi_1^2 + psi_2^2)
l_3


```

$$\begin{equation}
   (1 - .4B + .6B^2 - .8B^3)(X_t - 9) = a_t
\end{equation}$$


```{r 7.6.7 message=FALSE}

# Assume that the variance of the error term is **3.49** and you are trying to find the prediction limits for the following model.

# (1 - .4B + .6B^2 -.8B^3)(X_t -9) = a_t
  
# 1.  What is psi_3 for this model?  (Round to the nearest thousandth)

# 2.  What is sigma_hat_a  (standard deviation rather than variance)  Round to the nearest hundredth

# 3.  What is the half-width for the upper 95th% prediction limit for the 3rd prediction (l = 3)?  Round to the nearest hundredth


psi.weights.wge(phi = c(.4, -.6, .8), lag.max = 5)

sigma_hat = sqrt(3.49)
sigma_hat

psi_0 = 1
psi_1 = .400
psi_2 = -.440
psi_3 = -.384

l_1 = (1.96) * (sigma_hat) * psi_0
l_1

l_2 = (1.96) * (sigma_hat) * sqrt(1 + psi_1^2)
l_2

l_3 = (1.96) * (sigma_hat) * sqrt(1 + psi_1^2 + psi_2^2)
l_3


```

```{r 7.6.8.1 message=FALSE}


# AR(1) Phi Positive
# data(fig6.1nf)
# plotts.wge(fig6.1nf)

fore.arma.wge(fig6.1nf, phi = .8, n.ahead = 20, plot = TRUE, limits = TRUE)

```

```{r 7.6.8.2 message=FALSE}


# AR(2, 1) 
data(fig6.2nf)

fore.arma.wge(fig6.2nf, phi = c(1.2, -.6), theta = .5, n.ahead = 20, plot = TRUE, limits = TRUE)

```

## 7.7 Checking Forecasts: Cross Validation


### 7.7.2 LASTN and ASE in TSWGE

ASE = Average Squared Error


```{r 7.7.2.1 message=FALSE}


data("llynx")
plotts.wge(llynx)

lynx_forecast_ar4 = fore.arma.wge(llynx, phi = c(1.3, -0.7, 0.1, -0.2), n.ahead = 30, limits = FALSE, lastn = TRUE)
ase = mean((lynx_forecast_ar4$f - llynx[85:114])^2)
ase

```


```{r 7.7.2.2 message=FALSE}


lynx_forecast_ar11 = fore.arma.wge(llynx, phi = c(1.17, -0.54, 0.27, -0.31, 0.15, -0.15, 0.06, -0.03, 0.13, 0.20, -0.34), n.ahead = 30, limits = FALSE, lastn = TRUE)
ase = mean((lynx_forecast_ar11$f - llynx[85:114])^2)
ase

```
### Soutwest Airlines Data

```{r 7.7.3.1 message=FALSE}

df_swa <- read.csv(file="../SWADelay.csv", header = TRUE)

swa_forecast_ar12 = fore.arma.wge(df_swa$arr_delay, phi = c(.44, .02, -.12, .08, .00, .02, .06, -.09, .06, .07, .02, .37), n.ahead = 30, limits = FALSE, lastn = TRUE)

ase = mean((swa_forecast_ar12$f - df_swa$arr_delay[148: 177])^2)
ase

```

```{r 7.7.3.2 message=FALSE}

# df_swa <- read.csv(file="../SWADelay.csv", header = TRUE)

swa_forecast_arma121 = fore.arma.wge(df_swa$arr_delay, phi = c(.34, .07, -.11, .07, .01, .02, .06, -.09, .05, .07, .02, .39), theta = c(-.12), n.ahead = 30, limits = FALSE, lastn = TRUE)

ase = mean((swa_forecast_arma121$f - df_swa$arr_delay[148: 177])^2)
ase

```

## 7.8 Forecasting with ARIMA


### ARIMA(0,1,0) Example

```{r 7.8.3.1 echo=FALSE, message=FALSE, results='hide'}

xd1 = gen.aruma.wge(n=75, d=1, sn=74)
fore.aruma.wge(xd1, d = 1, n.ahead = 5, limits = FALSE)

```
### ARIMA(0,2,0)

```{r 7.8.3.1 echo=FALSE, message=FALSE, results='hide'}

xd1 = gen.aruma.wge(n=30, d=2)
fore.aruma.wge(xd1, d = 2, n.ahead = 5, limits = FALSE)

```


### More ARIMA Examples

#### ARIMA(0,1,0)

```{r 7.8.4.1 echo=FALSE, message=FALSE, results='hide'}

xd1 = gen.aruma.wge(n=50, phi = .8, d=1, sn=15) # ARIMA (1,1,0)
fore.aruma.wge(xd1, d = 1, n.ahead = 20, limits = FALSE) # Forecast (0,1,0)

```

#### ARIMA(2,1,0)

```{r 7.8.4.1 echo=FALSE, message=FALSE, results='hide'}

xd1 = gen.aruma.wge(n=50, phi = .8, d=1, sn=15) # ARIMA (1,1,0)
fore.aruma.wge(xd1, phi = .8, d = 1, n.ahead = 20, limits = FALSE) # Forecast (0,1,0)

```

#### ARIMA(0,2,0)

```{r 7.8.4.1 echo=FALSE, message=FALSE, results='hide'}

xd1 = gen.aruma.wge(n=50, phi = .8, d=1, sn=15) # ARIMA (1,1,0)
fore.aruma.wge(xd1, d = 2, n.ahead = 20, limits = FALSE) # Forecast (0,1,0)

```


## 7.9 Forecasting with Seasonal Models

### TSWGE

Generating and Forecasting Simple Seasonal Models

```{r 7.9.1 echo=FALSE, message=FALSE}

xs4 = gen.aruma.wge(n=20, s=4, sn=6) 
fore.aruma.wge(xs4, s = 4, n.ahead = 8, limits = FALSE) 

fore.aruma.wge(xs4, s = 4, n.ahead = 8, limits = FALSE, lastn = TRUE) 

```


```{r 7.9.2 echo=TRUE, message=FALSE}

xs4 = gen.aruma.wge(n=20, phi = .8, s=4, sn=6) 
fore.aruma.wge(xs4, phi = .8, s = 4, n.ahead = 24, limits = FALSE) 

```

### Airlog Data

#### Parzen

```{r 7.9.4.1 echo=TRUE, message=FALSE}

data(airlog)

parzen = fore.aruma.wge(airlog, d = 0, s = 12, phi = c(.74, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, .38, -.2812), n.ahead = 36, lastn = TRUE, limits = FALSE)

parzen_ase = mean((airlog[(144 - 36 + 1): 144] - parzen$f)^2)
parzen_ase

```



## 7.9 Forecasting with Seasonal Models

#### Box

```{r 7.9.4.2 echo=FALSE, message=FALSE, results='hide'}

# data(airlog)

box = fore.aruma.wge(airlog, d = 1, s = 12, theta = c(.4, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, .6, -.24), n.ahead = 36, lastn = TRUE, limits = FALSE)

box_ase = mean((airlog[(144 - 36 + 1): 144] - box$f)^2)
box_ase

```



#### Woodward/ Gray

```{r 7.9.4.2 echo=FALSE, message=FALSE, results='hide'}

# data(airlog)

wood = fore.aruma.wge(airlog, d = 1, s = 12, phi = c(-.36, -.05, -.14, -.11, .04, .09, -.02, .17, .03, -.10, -.38), n.ahead = 36, lastn = TRUE, limits = FALSE)

wood_ase = mean((airlog[(144 - 36 + 1): 144] - wood$f)^2)
wood_ase

```



## 7.10 Signal Plus Noise Forecasts

### TSWGE


```{r 7.10.1 echo=FALSE, message=FALSE, results='hide'}

x = gen.sigplusnoise.wge(n = 50, b0 = 10, b1 = .2, phi = c(.8, -.6))

xfore = fore.sigplusnoise.wge(x, linear = TRUE, freq = 0, max.p = 5, n.ahead = 10, lastn = FALSE, plot = TRUE, limits = FALSE)


```


### Amazon Stock Price 

```{r 7.10.1 echo=FALSE, message=FALSE, results='hide'}

df_amazon <- read.csv(file="../AMZN.csv", header = TRUE)

# x = gen.sigplusnoise.wge(n = 50, b0 = 10, b1 = .2, phi = c(.8, -.6))
# 
xfore = fore.sigplusnoise.wge(df_amazon$Adj.Close, linear = TRUE, freq = 0, max.p = 5, n.ahead = 30, lastn = FALSE, plot = TRUE, limits = TRUE)


xfore$f[30]

max(xfore$ul)

min(xfore$ll) # 1703.74
max(xfore$ll)

```


```{r}
# Date Manipulation
library(lubridate)

# Metrics
library(MLmetrics)

# Time Series Analysis
library(tseries)
library(forecast)
library(TSstudio)

```


```{r}

df_optiplex <- read.csv(file="./data/sc_req_item.csv", header = TRUE)


# df_optiplex$opened_at = as.POSIXct(strptime(df_optiplex$opened_at, "%Y-%m-%d %H:%M:%SS"))

df_optiplex$opened = floor_date(mdy_hm(df_optiplex$opened_at), "day")


```

```{r}
# ggplot(data=df_optiplex, aes(x = opened, y = quantity)) +
#       geom_line() +
#       ggtitle("Monthly Optiplex Purchases") +
#       labs(x = "Curve Day", y = "Number of Cases")

# https://stackoverflow.com/questions/33221425/how-do-i-group-my-date-variable-into-month-year-in-r/33221885

gb <- df_optiplex %>% 
    group_by(month = lubridate::floor_date(opened, "month")) %>%
    summarize(summary_variable = sum(quantity))

```

```{r}

ggplot(data=gb, aes(x = month, y = summary_variable)) +
      geom_line() +
      ggtitle("Monthly Optiplex Purchases") +
      labs(x = "Month", y = "Quantity Purchased")

```



```{r}

arima1 <- arima(gb$summary_variable, order=c(1, 1, 1))

```

```{r}
forcast1 <- forecast(arima1, h=3)
```

```{r}
plot(forcast1)
legend(x = "topleft", legend = c("Predicted"), col = c("blue"), lty = c(1, 1))
```

