---
title: "R Notebook"
---


```{r library-imports}

library(tidyverse)
library(tswge)

```


# Unit 1 - Stationarity

## 1.2 Notation, Terminology, Motivation, and Examples



## 1.3 Review of Expected Value, Variance, Covariance, and Correlation

### 1.3.1 Expected Value

$$\begin{equation}
\mathrm{E}[X] = \mu
\end{equation}$$

## 1.3.2 Variance


$$\begin{equation}
Var(X) = \mathrm{E}[(X - \mu)^2]
\end{equation}$$


$$\begin{equation}
   \hat{\gamma_0} = \hat{\sigma^2_x} = \frac{\sum(x_i - \bar{x})^2}{n}
\end{equation}$$


## 1.3.3 Covariance



## 1.4 Stationarity: Condition 1

Subpopulations of X_t have the same mean for each t. Restated, the mean does not depend on time (t).

## 1.5 Stationarity: Condition 2

Subpopulations of X for a given time have a finite and constant variance for all t. Restated, the variance does not depend on time.


## 1.6 Stationarity: Condition 3, Part I

## 1.7 Stationarity: Condition 3, Part II

The correlation of X_t1 and X_t2 depends only on t_2 - t_1. That is,the correlation between 
data points is dependent only on how far apart they are, not where they are. 

## 1.8 Stationary Time Series

## 1.9 Estimation, Part I

## 1.10 Estimation, Part II

## Howework



# Unit 4 - AR(2) and AR(p)

## 4.1 Review: AR(1)

 - Realizations are wandering or oscillating depending on whether the root of the characteristic equation is positive or negative, respectively.

- Autocorrelations are damped exponentials depending on whether the root of the characteristic equations is positive or negative, respectively.

- Spectral densities have a peak at f=0 or f=.5 depending on whether the root of the characteristic equations is positive or negative, respectively.


## 4.2 AR(2): Introduction and Stationarity


### 4.2.2

Expected value

$$\begin{equation}
\mathrm{E}[X_t] = \mu
\end{equation}$$



Variance

$$\begin{equation}
   \sigma_X^2 = \frac{\sigma_a^2}{1 - \phi_1\rho_1 - \phi_2\rho_2}
\end{equation}$$



### 4.2.3  AR(2) Processes: Zero Mean Form

Alternative form of zero mean AR(2) model

$$\begin{equation}
   X_t - \phi_1X_{t-1} - \phi_2X_{t-2} = a_t
\end{equation}$$

## 4.3 AR(2): Backshift Operator


$$\begin{equation}
   X_t - \phi_1BX_t - \phi_2B^2X_t = a_t
\end{equation}$$


or 


$$\begin{equation}
   (1 - \phi_1B - \phi_2B^2)X_t = a_t
\end{equation}$$

## 4.4 AR(2): Characteristic Equation and Stationarity

### 4.4.1

An AR(2) model is stationary if and only if the roots of the characteristic equation are outside of the unit circle.


Characteristic Equation

$$\begin{equation}
   1 - \phi_1Z - \phi_2Z^2 = 0
\end{equation}$$


Roots of the characteristic equation

1.  Either 2 real roots
2.  r1, r2 will be complex conjugate pairs


```{r 4.4.7}

# X_t - .4X_t-1 + .7X_t-2 = a_t



a = -.7
b = .4
c = 1


z = ( -b + sqrt(b^2 - (4 * a * c)) ) / (2 * a)
z

z = ( -b - sqrt(b^2 - (4 * a * c)) ) / (2 * a)
z

# Factor Table
factor.wge(phi = c(.4, -.7))


# plotts.true.wge(phi = c(.4, -.7))


```



## 4.5 AR(2): Behavior—Case 1: Two Real Roots


### 4.5.2


```{r 4.5.2 }

# Xt - .2X_t-1 - .48X_t-2  Two Positive


factor.wge(phi = c(-.9, -.2))

# x = gen.arma.wge(200, phi = c(.2, .48))
# # 
# plotts.sample.wge(x)

```

# Unit 5 | Moving Average (MA(q)) and ARMA(p,q) Models

## 5.2 MA(q): Properties and Characteristics, Part I

### Equation for rho 1

$$\begin{equation}
   \rho_1 = \frac{-\theta_1}{1 + \theta_1^2}
\end{equation}$$



```{r 5.2.2}

# X_t = a_t -.99a_t-1 


theta1 = .9

p0 = 1
p1 = NA
p2 = 0
p3 = 0


p1 = (-(theta1)) / (1 + (theta1^2))
p1



```

### Equation for rho 2

$$\begin{equation}
   \rho_2 = \frac{-\theta_2}{1 + \theta_1^2 + \theta_2^2}
\end{equation}$$



```{r 5.2.7}

# X_t = a_t -.5a_t-1 + .9a_t-2


theta1 = .5
theta2 = -.9

p0 = 1
p1 = NA
p2 = NA
p3 = 0


p1 = (-(theta1) + (theta1 * theta2)) / (1 + (theta1^2 + theta2^2))
p1


p2 = -(theta2) / (1 + (theta1^2 + theta2^2))
p2



```


## 5.3 MA(q), Properties and Characteristics, Part II

MA behavior is most useful in the formulation of ARMA(p, q) models.

### 5.3.1 tswge


```{r 5.3.1.1}

# Generating MA(1) data

# gen.arma.wge(n=100, theta=.99)
# gen.arma.wge(n=100, theta=-.99)
# gen.arma.wge(n=100, theta=.99, sn=5)
# gen.arma.wge(n=100, theta=-.99, sn=5)
plotts.true.wge(theta=c(-.1, .3))

```


## 5.4 MA(q), Invertibility


Note:  Two models with the same autocorrelations is undesirable.  (model multiplicity)


MA models are __invertible__ if and only if all roots of the MA characteristic equation are outside of the unit circle.


Is this model invertible?  Yes, its root is outside the unit circle.

$$\begin{equation}
   X_{t} = a_{t} + .7X_{a-1}
\end{equation}$$



```{r}


# X_t = (1 + .7B)a_t
# 1 + .7Z = 0 (characteristic equation)

z = -1 / .7

z

```

### 5.4.3 MA(q) Invertablity

```{r 5.4.3.1}


# MA(2) X_t = a_t -1.6a_t-1 + .9a_t-2 

factor.wge(phi=c(1.6,-.9))

### Abs Recip < 1 (and therefore outside of the unit circle which is invertable)

```


## 5.5 ARMA(p,q), Properties and Characteristics

```{r 5.5.5}

# factor.wge(phi = c(-.1, -.82, .16,))

# plotts.true.wge(phi = c(-.1, -.82, .16), theta = -.2)

plotts.true.wge(phi = c(.1, -.5, .7), theta = c(.72, -.8))


```



# Unit 7 - Forecasting

## 7.2 Forecasting Is Extrapolating


## 7.3 Strategy and Notation

t_0 is called the forecast origin

l is the lead time, i.e, the number of time units (steps ahead) which we want to forecast

X_hat_10 (2) is the forecast of X_12

## 7.4 Forecasting with AR(p) Models

$$\begin{equation}
   {X_t} - \phi_1{X_t}_1 = (1 - \phi_1)\mu + a_t
\end{equation}$$

## 7.5 Forecasting with ARMA(p,q)

## 7.6 Probability Limits and ARMA Forecasts

### 7.6.4 Calculating Psi Weights with TSWGE

 - psi_0 = 1, psi_1 = .7, psi_2 = .24, psi_3 = -.132, ...

 - psi.weights.wge(phi = c(1.2, -.6), theta = .5, lag.max = 5)


## 7.7 Checking Forecasts: Cross Validation




```{r}


data(UKDriverDeaths)

aic5.wge(UKDriverDeaths)

```

```{r}

ar_4_2 = fore.arma.wge(UKDriverDeaths, phi = c(-0.67, 0.47, 0.36, -0.18), theta = c(-1.46, -0.61), n.ahead = 12, lastn = TRUE, limits = FALSE)


```

```{r}
ar_4_2_ase = mean((UKDriverDeaths[(192 - 12 + 1): 192] - ar_4_2$f)^2)
ar_4_2_ase
```



```{r}

seasonalf = fore.aruma.wge(UKDriverDeaths, s = 12, n.ahead = 12, lastn = FALSE, limits = TRUE)


```


```{r}

seasonal_ase = mean((UKDriverDeaths[(192 - 12 + 1): 192] - seasonal$f)^2)
seasonal_ase


```

```{r question-two-a}

# (1 - .9B)(1 - .8B)X_t = a_t

# ()

# factor.wge(c(.3, -.9))

psi.weights.wge(phi = c(.3, -.9), lag.max = 5)

```

```{r}


X_1 = 1
X_2 = 5
X_3 = 4
X_4 = 6
X_5 = 8
X_6 = 10
X_7 = 9
X_8 = 11

X_bar = c(1, 5, 4, 6, 8, 10, 9, 11)

sigma_hat = sqrt(1.82)

psi_0 = 1
psi_1 = 0.30000
psi_2 = -0.81000
psi_3 = NA


l_1 = (1.96) * (sigma_hat) * psi_0
l_1

l_2 = (1.96) * (sigma_hat) * sqrt(1 + psi_1^2)
l_2

l_3 = (1.96) * (sigma_hat) * sqrt(1 + psi_1^2 + psi_2^2)
l_3

```



```{r}

X = c(10.05, -9.42, 8.82, -8.34, 8.07, -7.76, 7.46, -7.22, 7.01, -6.66, 6.28, -5.87, 5.46, -4.97, 4.65, -4.30, 4.01, -3.72, 3.24, -2.81, 2.44)

mean(X)

summarise(X)

```

