---
title: "Homework 1"
output: html_notebook
---

```{r library-imports}

library(tidyverse)
library(tswge)

```


# Exercises

## Applied Problems

### 1.1  

The following data are annual sales of a hypothetical company in millions of dollars.

See 1.9 Estimation, Part I for examples.

```{r}
# xdf = read.csv(file.choose(), header=TRUE)

# x = as.numeric((paste(xdf$Adj.Close)))

x <- c(76, 70, 66, 60, 70, 72, 76, 80)

x <- x[!is.na(x)]

n <- length(x)  # n = 8

nlag <- n - 1 # n - 1 (7)


m <- mean(x)

v <- var(x, na.rm = TRUE)

gamma0 = var(x) * (n-1) / n

aut <- acf(x, lag.max = nlag) # n - 1 (1508)

# sum <- 0

# for (k in 1:nlag) {sum = sum + (1 - k/n) * aut$acf[k+1] * gamma0}
# vxbar = 2 * sum / n + gamma0 / n # note the multiplication of the sum by 2
# vxbar

sum_gamma_0 <- 0

for (k in 1:n) {sum_gamma_0 = sum_gamma_0 + (x[k] - m)^2}
gamma_0_manual <- sum_gamma_0 / n


sum_gamma_1 <- 0

for (k in 1:nlag) {sum_gamma_1 = sum_gamma_1 + (x[k] - m) * (x[k+1] - m)}
gamma_1_manual <- sum_gamma_1 / n



rho_1 <- gamma_1_manual / gamma_0_manual


```
Sigma^2 Hat = Gamma 0

$$\begin{equation}
\hat\gamma_0 = 34.9375
\end{equation}$$


Gamma 1

$$\begin{equation}
\hat\gamma_1 = 14.74219
\end{equation}$$

Rho Hat 0
$$\begin{equation}
\hat\rho_0 = 1
\end{equation}$$


Rho Hat 1
$$\begin{equation}
\hat\rho_1 = \hat\gamma_1 \div \hat\gamma_0 = .422
\end{equation}$$


### 1.2

The data shows a non-cyclic wandering behavior with strong correlation between values taht are near each other in time.  The sample auto correlations show strong positive correlation (above 0.5 for k <= 13 ), and the periodogram and Parzen spectral estimator show peaks at zero with no indication of cyclic behavior.



#### West Texas Intermediate Crude Oil Prices

```{r}
data(wtcrude)
plotts.sample.wge(wtcrude)
```
#### Pennsylvania average monthly temperatures

The data show a strong cyclic behavior with a period of about 12 which makes sense (because this is monthly data).  The sample auto correlations show a very slowly damping cyclic behavior with cycle length of 12 while the periodogram and Parzen spectral estimator show a strong peak at about f=1/12.

```{r}
data(patemp)
plotts.sample.wge(patemp)
```

### 1.4

Realization 1

Wandering behavior which corresponds to fairly high positive auto-correlations for lags of modest length (d) and has a spectral density with a peak at zero (a).

- Auto correlations (D)
- Spectral Density (A)

Realization 2

Has very little structure (random line) which corresponds to small or zero auto-correlations (a) and a flat spectral density (d)

- Auto correlations (A)
- Spectral Density (D)


Realization 3

Pseudo-cyclic with a period of about 10 which corresponds to the auto-correlations with damped sinusoidal behavior of about 10 (b) and a peak in the spectral density of about f=0.1 (c)


- Auto correlations (B)
- Spectral Density (C)

Realization 4

Seems to have a pseudo-cyclic behavior with period of a little less than 10 along with a higher frequency component.  This corresponds to the auto-correlations in (c) that show a damped sinusoidal behavior of about 2 along with some high-frequency distortion.  The spectral density in (b) shows a peak at about f=1/12 along with a higher frequency peak.

- Auto correlations (C)
- Spectral Density (B)



### 1.6


The realizations shows a dominant frequency with period of about 20 along with a higher-frequency behavior.  The auto correlations show the periodic behavior of associated with the period 20 (f=0.05) with some indication of a higher frequency component.  These two plots provide very little evidence regarding the nature of the higher frequency behavior.  The two spectral plots clearly suggest frequency behavior at both f=0.05 and f=0.35.

```{r}

x <- gen.sigplusnoise.wge(n=100, coef=c(3, 1.5), freq=c(.05, .35), psi=c(0,2))
plotts.sample.wge(x)

```

