---
title: "Unit 6 - ARIMA Models and Seasonality"
output: html_notebook
---

# For Live Session

1. Looking at your time series from the first day of class, you addressed its stationarity before. Does either a signal plus noise, ARIMA, or ARUMA (seasonal) model seem appropriate?  (Just comment on your PowerPoint slide. We will test the fits of these models later.)

2. Copy and paste the factor table for a seasonal model with s = 7. 

3. Consider the following models:

Model 1

$$\begin{equation}
   (1 - B^4)X_{t} = (1 - .6B + .8B^2)a_{t}
\end{equation}$$


Model 2

$$\begin{equation}
   (1 + .3B)X_{t} = (1 - .6B + .8B^2)a_{t}
\end{equation}$$

Model 3


$$\begin{equation}
   (1 - .1B - .99B^2 + .013B^3 + .2078B^4 + .0888B^5 + .00864B^6)(1 - B)^2(1 - B^{12})X_{t} = (1 - .6B + .8B^2)a_{t}
\end{equation}$$



Comment and provide evidence (for each of the models) if they look to be:

ARUMA with d ≠0 and s≠0
ARUMA with d = 0 and s≠0 (Identify p and q as well.)
ARIMA (Identify d as well as p and q.)
ARMA (Identify p and q.)



4. Pick a stock and download the last year's worth of daily stock price data.  

- Plot the data. 
- Take the first difference, and then estimate the structure of differenced data.  
- From this information, suggest a model for the data. 

5. Provide a slide or slides for your Key Takeaways for Unit 6!

6. Provide a slide or slides for any questions or comments you have from this section. This could also include topics you would like to have specifically addressed in live session! There is no minimum or maximum here.


```{r library-imports}

library(tidyverse)
library(tswge)

```


## Bikeshare Demand

```{r for-live-session-google-doc-1}

df_train <- read.csv(file="../../Unit 2/pre_live/data/train.csv", header = TRUE)

# df_train$datetime = as.POSIXct(df_train$datetime)

df_train$datetime = as.POSIXct(strptime(df_train$datetime, "%Y-%m-%d %H:%M:%S"))


df_train <- df_train %>%
  mutate(year = as.factor(format(datetime, format = "%Y")), 
         month = as.numeric(format(datetime, format = "%m")), 
         day = as.factor(format(datetime, format = "%d")),
         hour = as.factor(format(datetime, format = "%H")))

```


```{r for-live-session-google-doc-2}

df_dec_2012 <- df_train %>%
    filter(datetime >= "2012-12-01")
  
```

```{r for-live-session-google-doc-3}

aic5.wge(df_dec_2012$count)

```






# Asynchronous Material

## 6.2 Signal Plus Noise Models vs. Stationary Models


### 6.2.4 Generate data from a signal-plus-noise model


```{r 6.2.4}

# Randomly generating signal plus noise models

gen.sigplusnoise.wge(100, b0 = 2, b1 = 4, vara = 100)
gen.sigplusnoise.wge(100, b0 = 0, b1 = 0, vara = 10)
gen.sigplusnoise.wge(100, b0 = 0, b1 = 0, phi = .975, vara = 10)

# Periodic Signal

gen.sigplusnoise.wge(100, coef = c(5, 0), freq = c(.1, 0), psi = c(.25, 0), phi = .975, vara = 10)

# AR(4) from the slides


# mult.wge 

# "The function multiplies the AR (or MA) factors of a model to produce the model in unfactored form. Requires the CRAN package 'PolynomF'."

parms = mult.wge(c(.975), c(.2, -.45), c(-.53))
parms$model.coef

gen.arma.wge(160, phi = parms$model.coef, vara = 1)



```

## 6.3 ARIMA Models, Part I

#### ARIMA Properties and Characteristics

ARIMA(p,d,q) Models

autoregressive integrated moving average process of orders p, d and q.

$$\begin{equation}
   \phi(B)(1 - B)^dX_t = \theta(B)a_{t}
\end{equation}$$

Generate realizations from an ARIMA model.  **gen.arima.wge**

### 6.3.3 P_k for an ARIMA process

$$\begin{equation}
   \sigma^2_x = \gamma_0 = \frac{\sigma^2_a}{ 1 - \phi^2_1 }
\end{equation}$$

```{r 6.3.3}

variance = 2
# phi_1 = .4
phi_1 = .99

sigma_squared_x  = (variance) / (1 - phi_1^2)
sigma_squared_x

```
### 6.3.4 ARIMA(0,1,0)

- Stricly wandering since there is one factor with a root of 1
- Identically equal to p_k = 1
- Slowly damping, even though the true values don't damp

$$\begin{equation}
   (1 - B)X_t = \alpha_t
\end{equation}$$



```{r 6.3.4}

x = gen.arima.wge(200, phi = 0, var = 1, d = 1, sn = 31)
acf(x)

```



## 6.4 ARIMA Models, Part II

### 6.4.1


```{r 6.4.1}

# the following command differences the data in x

# y = artrans.wge(x, phi.tr = 1)

# This simply means that y(i) = x(i) - x(i - 1)

# y has length n-1 because x(1) has no x(0) before it.

# Example

x = c(1, 3, 6, 10, 25)

y = artrans.wge(x, phi.tr = 1)

y # shows the 4 differences



```

### ARIMA(0,1,0)

```{r 6.4.2.1}

# R Code

# (1 - B)X_t = a_t

x = gen.arima.wge(200, phi = 0, var = 1, d = 1, sn = 31)
acf(x)

Xtilde = artrans.wge(x, 1)
plotts.wge(Xtilde)
aic5.wge(Xtilde)

```

### ARIMA(2,1,0)


```{r 6.4.2.2}

# (1 - 1.5B + .8B^2)(1 - B)X_t = a_t

# R Code

a = gen.arima.wge(200, phi = c(1.5, -.8), var = 1, d = 1, sn = 31)
acf(a)

model = mult.wge(fac1 = c(1.5, -.8), fac2 = 1)
factor.wge(model$model.coef)

FirstDif = artrans.wge(a, 1) # Take out the (1-B)
parzen.wge(FirstDif)
aic5.wge(FirstDif) # Check the structure of the noise

```


### ARIMA(2,2,1)

```{r 6.4.2.3}

# (1 - 1.5B + .8B^2)**(1 - B)^2**X_t = (1 + .8B)a_t

# R Code

a = gen.arima.wge(200, phi = c(1.5, -.8), theta = -.8, var = 1, d = 2, sn = 21)
acf(a)


p = parzen.wge(a, trunc = 40)
ar = mult.wge(fac1 = c(1.5, -.8), fac2 = 1, fac3 = 1)
factor.wge(ar$model.coef)


FirstDif = artrans.wge(a, 1) # Take out the (1-B)

SecondDif = artrans.wge(FirstDif, 1) # Take out the other (1-B)

parzen.wge(SecondDif)

aic5.wge(SecondDif) # Check the structure of the noise



```



### ARIMA(2,2,1)

```{r 6.4.3}

# (1 - .6B + .8B^2)(1 - B)^2X_t = (1 + .3B)a_t

# R Code

a = gen.arima.wge(500, phi = c(.6, -.8), theta = -.3, var = 1, d = 2, sn = 35)
acf(a)


p = parzen.wge(a, trunc = 40)
ar = mult.wge(fac1 = c(.6, -.8), fac2 = 1, fac3 = 1)
factor.wge(ar$model.coef)


FirstDif = artrans.wge(a, 1) # Take out the (1-B)

SecondDif = artrans.wge(FirstDif, 1) # Take out the other (1-B)

parzen.wge(SecondDif)

aic5.wge(SecondDif) # Check the structure of the noise



```